Transition path sampling based calculations of free energies for enzymatic reactions: the case of human methionine adenosyl transferase and plasmodium vivax adenosine deaminase

### 2022-05-10 tex/manuscript.tex
final corrections

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Meta-data block
%% ---------------
%% Each author should be given as a separate \author command.
%%
%% Corresponding authors should have an e-mail given after the author
%% name as an \email command. Phone and fax numbers can be given
%% using \phone and \fax, respectively; this information is optional.
%%
%% The affiliation of authors is given after the authors; each
%% \affiliation command applies to all preceding authors not already
%% assigned an affiliation.
%%
%% The affiliation takes an option argument for the short name.  This
%% will typically be something like "University of Somewhere".
%%
%% The \altaffiliation macro should be used for new address, etc.
%% On the other hand, \alsoaffiliation is used on a per author basis
%% when authors are associated with multiple institutions.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\author{Sree ganesh balasubramani}
\affiliation{Department of Chemistry and Biochemistry, University of Arizona, Tucson, Arizona 85721, United States}
\author{Steven D. Schwartz}
\affiliation{Department of Chemistry and Biochemistry, University of Arizona, Tucson, Arizona 85721, United States}
%\author{I. Ken Groupleader}
%\altaffiliation{A shared footnote}
\email{sschwartz@email.arizona.edu}
%\phone{+123 (0)123 4445556}
%\fax{+123 (0)123 4445557}
%\affiliation[Unknown University]
%{Department of Chemistry, Unknown University, Unknown Town}
%\alsoaffiliation[Second University]
%{Department of Chemistry, Second University, Nearby Town}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% The document title should be given as usual. Some journals require
%% a running title from the author: this should be supplied as an
%% optional argument to \title.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\title[]
  {Transition path sampling based calculations of free energies of enzymatic
  reactions: the case of human methionine adenosyl transferase and plasmodium 
  vivax adenosine deaminase}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Some journals require a list of abbreviations or keywords to be
%% supplied. These should be set up here, and will be printed after
%% the title and author information, if needed.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\abbreviations{TPS,WHAM}
\keywords{American Chemical Society, \LaTeX}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% The abstract environment will automatically gobble the contents
%% if an abstract is not used by the target journal.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{abstract}
  Transition path sampling (TPS) is widely used for the
  calculations of reaction rates, transition state structures and reaction 
  coordinates of condensed phase systems. Here we discuss a scheme 
  for the calculation of free energies using the ensemble of TPS reactive 
  trajectories in combination with a window based sampling technique 
  %introduced by 
  %Radhakrishnan et al. (Radhakrishnan, R.; Schlick, T., J. Chem. Phys. 2004, 121, 2436.) 
  for enzyme catalyzed reactions. We calculate the free energy profiles of 
  the reactions catalyzed by the human methionine S-adenosyltransferase (MAT2A) 
  enzyme and the plasmodium vivax adenosine deaminase enzyme to assess 
  the accuracy of this method. MAT2A catalyzes the formation of S-adenosine-L-methionine
  following a S$_{\text{N}}2$ mechanism and using our method we estimate
  the free energy barrier to be $16\;\text{Kcal}\;\text{mol}^{-1}$ which
  is in excellent agreement with the experimentally measured activation 
  energy of $17.27\;\text{Kcal}\;\text{mol}^{-1}$. Furthermore for the 
  $pv$ADA enzyme catalyzed reaction we estimate a free energy barrier of 
  $23\;\text{Kcal}\;\text{mol}^{-1}$ and the calcualted free energy profile is
  similar to that predicted from experimental observations.
  We show that calculating free energies within TPS is lucrative as it is 
  bias free, accurate compared to experimental measurements 
  and computationally reasonable.    
\end{abstract}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Start the main part of the manuscript here.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Introduction}
Molecular dynamics (MD) simulations are increasingly being
used to calculate quantitative estimations of experimentally measurable 
kinetic and thermodynamic parameters of enzyme catalyzed reactions. 
\cite{Karplus02NatStructMolBiol9p646,
Schramm98AnnuRevBiochem67p693,Zhang05AccChemRes38p379,
Schramm11AnnuRevBiochem80p703,Schramm18ChemRev118p11194} 
This has helped in proposing mechanisms and explanations for 
enzyme activity, fast high throughput screening of thousands of inhibitor 
molecues in drug lead discovery, \cite{Jorgensen09AccChemRes42p724,Sliwoski14PharmacolRev66p334} 
directed evolution for designing enzymes with enhanced catalytic 
activities, \cite{Thyme09Nature461p1300,Bloom09PNAS106p9995,Schafer19JAmChemSoc141p10431} etc.    
Straightforward MD simulations adequately sample 
the long-lived stable reactant and product states 
but the high energy barrier-crossing events that occur rarely and at long 
time scales are not easily accessible using this method alone.
To access such states at a higher frequency using MD simulations 
several enhanced sampling techniques 
have been developed in the past few decades some of which are 
umbrella sampling, \cite{Kastner11WileyInterdiscipRevComputMolSci1p932} 
metadynamics, \cite{Barducci11WileyInterdiscipRevComputMolSci1p826}
steered MD, \cite{Park04JChemPhys120p5946}
milestoning, \cite{Faradjian04JChemPhys120p10880} and transition path sampling 
(TPS). \cite{Pratt86JChemPhys85p5045,Bolhuis02AnnRevPhysChem53p291,
Dellago98JChemPhys108p1964,Bolhuis21AdvTheory4p2000237}

TPS in particular is an attractive method since it samples reactive trajectories
in which the rare barrier-crossing events are guaranteed to occur and
it does not require a priori knowledge of the reaction coordinate which can be 
complex for large biological systems.
TPS is based on the Monte Carlo importance sampling in the 
space of trajectories. It uses the shooting and shifting algorithm 
to collect an ensemble of reactive pathways that are real dynamical 
trajectories generated without the application of any bias forces. \cite{dellago02AdvChemPhys123} 
Therefore the TPS ensemble can be used to obtain 
kinetic information such as the rate of the reaction as well as to elucidate 
the transition state, reaction coordinate and the reaction mechanism. 
TPS has been successfully applied to study problems such as enzyme catalysis, 
\cite{Antoniou01JPhysChemB105p5553,Hay12NatChem4p161,Schwartz09NatChemBiol5p551} 
protein folding, \cite{Bolhuis03ProcNatlAcadSci100p12129,Juraszek12ChemPhys396p30} etc.

Calculating equilibrium properties such as the free energy of an enzymatic 
catalysis reaction within TPS is not straighforward. 
The Monte Carlo algorithm used in TPS has an acceptance criterion that 
is only satisfied if the trajectory begins in the reactant basin and ends in 
the product basin or vice versa, hence trajectories that are localized in the 
reactant or product basins are rejected. The localized trajectories contribute
to the probability density distribution function ($P(\lambda)$) which is the 
probability of a system at equilibrium to have a particular value of the 
order parameter $\lambda$. \cite{Dellago09AdvCompSimAppp167}

Frenkel has shown that inclusion of rejected states in Markov state Monte 
Carlo algorith can help speed up the convergence of the calculations of 
order-parameter distribution of a two-dimensional
Ising model. \cite{Frenkel04ProcNatAcadSci101p17571}
Inclusion of rejected trajectories within TPS was first discussed by Radhakrishnan 
et al. \cite{Radhakrishnan04JChemPhys121p2436} who developed `BOLAS' which combines 
the shooting and shifting algorithms of TPS with a window based sampling technique 
in the spirit of umbrella sampling to calculate equilibrium free energy 
differences. Peters et al. used a variant of the BOLAS scheme called the 
equilibrium path sampling (EPS) where they use aimless shooting 
combined with window based sampling for free energy calculations. 
\cite{Peters08JAmChemSoc130p17342,Beckham10epsbook} 
More recently Brotzakis et al., \cite{Brotzakis19JChemPhys151p174111}
have developed a reweighted path ensemble scheme within the standard TPS method 
called virtual interface exchange TPS (VIE-TPS) to calculate the free energy 
landscape. VIE-TPS is shown to produce reasonably accurate free energy profiles for 
chemical reactions that are diffusive in nature whereas it is less accurate otherwise.
Transition interface sampling (TIS) has been used to obtain 
reweighted path ensemble which can give access to free energy 
and committor surfaces. \cite{Rogal10JChemPhys17p174109}

@@

distributed to almost all body tissues and fluids, is an universal 

@@

plasmodium vivax is a parasite that is responsible for the most number 

@@

is a good anti-malarial drug development strategy. \cite{Madrid08JBiolChem283p35899} 

@@

to calculate the free energies briefly, then we show calculations which apply this 
method on the two enzymatic reactions. Finally we compare our results to expeiments

@@

introducing notations as well as for the sake of being complete. 

@@

on $z_{i\Delta t}$ and not on conditions before the time $i\Delta t$ the total path probablility can 

@@

Within the transition path sampling, the trajectories of interest start in the reactant region of the 
phase space ($\mathcal{A}$) and end in the product region ($\mathcal{B}$). Such reactive trajectories have 
a restricted probability distribution function given by
\begin{equation}
\mathcal{P}_{\mathcal{AB}}[z_{\mathcal{T}}] = h_{\mathcal{A}}(z_0)\mathcal{P}[z(\mathcal{T})]
h_{\mathcal{B}}(z_{\mathcal{T}})/Z_{\mathcal{AB}}(\mathcal{T})\label{eqn:tpsensem}
\end{equation}
where 
\[
    h_{\mathcal{A}/\mathcal{B}}(z)= 
\begin{cases}
    1, & \text{if } z\in \mathcal{A}/\mathcal{B}\\
    0,              & \text{otherwise}
\end{cases}
\]
The set of all reactive trajectories characterized by the distribution function
given by Eq. \ref{eqn:tpsensem} is the transition path ensemble.

@@

The free energy as a function of the order parameter $\zeta$ is defined
as 

@@

where $P(\zeta)$ is the probability distribution of the reaction coordinate
$\zeta$ and $C$ is an arbitrary constant. At equilibrium the probability distribution of the 
reaction coordinate 
can be obtained from the distribution function for the phase space $\rho(\textbf{q})$ as

@@

The integration is over the entire phase space. In practice the calculation of this distribution function

@@

divided into bins and the frequency of the occurence of the reaction coordinate within a particular window
during the course of molecular dynamics simulations is used to calculate the probability distribution. 
Since some values of the reaction coordinates particulary close to the transition state are 

@@

without the application of any external bias, but the trajectories sampled with TPS are only reactive and hence

@@

The restriction that the pathways start from the reactant state and end in the product state can be relaxed
to obtain an ensemble of trajectories that will be distributed according to the equilibrium distribution. 

@@

\item Divide the order parameter into windows $\{\zeta_i\}$ within which the 
order parameters are given by 
$\zeta_{i}^{min} < \zeta < \zeta_{i}^{max}$.

@@

$\tilde{z}(\mathcal{T})$ and accept the new trajectory if at any time slice the calculated 

@@

accepted trajectories for the window $i$. 
\item Combine the probability distributions in successive windows by adjusting the constants $C$
such that the free energy $A$ from Eq. \ref{eqn:fenergy} is continuous. 

@@

The energies of both the systems were minimized using 50 steps of the 
steepest descent method, followed by 2000 steps of the
adopted basis Newton-Raphson method where only classical molecular mechanics 
was used for the dynamics. 
The minimized systems were
heated slowly to 300 K for 35 ps beginning with harmonic
constraints on all atoms except on the H atoms and the TIP3P 
water molecules with gradual reduction of the restraint forces to 1. 
15 ps of equilibration was carried out starting with harmonic 
constraints of force constant 1 followed by 20 ps of contraint free 
equilibration to prepare the systems for TPS simulations. 
During the heating and equilibration steps, bonds that contain 
hydrogen atoms in the MM region were restrained to their equilibrium values 
using the SHAKE procedure. \cite{Ryckaert77JComputPhys23p327} 
%------------------------------------------------------------------------------ 
\subsection{Transition path sampling}
%------------------------------------------------------------------------------
TPS simulations require the definition of unique reactant and product states
to collect reactive trajectories based on the Monte Carlo algorithm. Here we 
discuss the details of the reactions involving the MAT2A and the $pv$ADA enzymes 
to identify and define the order parameters characterizing the reactant and product basins.

The reaction catalyzed by the human MAT2A enzyme is
between adenosyl $5'$-triphosphate (ATP) and L-methionine (MET) molecules 
resulting in the formation of S-adenosyl-L-methionine (SAM) which is 
depicted in Fig. \ref{fig:mat2a-reaction}. This reaction follows a S$_{\text{N}}2$ 
mechanism and was experimentally characterized by 
Firestone et al. \cite{Firestone17JAmChemSoc139p13754}. The key bond 
parameters for this reaction consists of the distance between the
nucleophilic sulfur atom of MET and the $5'-C$ atom of the ATP (denoted as d$_{\mathrm{SC}}$), 
and the distance between oxygen atom of the phosphate group and the $5'-C$ 
atom of ATP (d$_{\mathrm{OC}}$). For the TPS calculations we define the 
reactant state to have d$_{\mathrm{OC}}$ to be less than 1.7 {\AA} and the product 
state to have the d$_{\mathrm{SC}}$ less than 1.7 {\AA}.

\begin{figure}
\includegraphics[width=0.5\textwidth]{figures/mat2a-reaction.png}
\caption{The reaction catalyzed by the human MAT2A enzyme between ATP and L-methionine resulting in the 
formation of S-adenosyl L-methionine and triphosphate leaving group.}
\label{fig:mat2a-reaction}
\end{figure}

$pv$ADA catalyzes the reaction between adenosine and OH$^{-}$ ion 
resulting in the formation of inosine and ammonia as shown in
Fig. \ref{fig:ada-reaction}.
The nucleophilic O atom of the OH$^{-}$ anion
attacks the electrophilic C6 atom of the adenosyl ring leading to the 
formation of a Meisenheimer complex like structure while the N1 atom 
on the adenosyl moiety gets protonated by the Glu229 residue of $pv$ADA. 
The key bond parameters are the distance between the approaching hydroxyl 
oxygen and the C6 atom (denoted as d$_{\text{OC}}$) and the 
dissociating C6$-$N6 (d$_{\text{NC}}$) bond distance. For the TPS simulaions 
we define the reactant state to have d$_{\text{NC}}$ being less than $1.7$ {\AA} 
and the product state to have the d$_{\text{OC}}$ less than $1.7$ {\AA}.

\begin{figure}
\includegraphics[scale=0.6]{figures/ada-new.png}
\caption{The reaction catalyzed by the $pv$ADA enzyme between adenosine and hydroxide ion
resulting in the formation of inosine and ammonia.}
\label{fig:ada-reaction}
\end{figure}

@@

250 steps forawards and backwards using $1\;fs$ time steps

@@

The order paratemer for the MAT2A system 

@@

is used to guide the window based umbrella sampling simulations. 
To sample in a particular window $w$, a frame from this reactive 

@@

order parameter falls within $w$. Using the shooting algorithm

@@

\caption{Equilibriated structure of the MAT2A enzyme in complex with ATP and 

@@

repeating this procedure several times and finding the fraction of trajectories that go 

@@

promote the reaction. Taking this into account the committor distribution analysis
starts by constraining the residues and molecules that are hypothesized to be part of the 

@@

\text{Kcal mol}^{-1}$ which is small Compared to typical enzyme catalyzed 
reactions.

@@

\caption{The free energy as a function of the order parameter.}

@@

\caption{Equilibriated structure of ADA protein}

@@

The approaching O atom of the OH$^{-}$ anion is at a distance of $1.48$ {\AA}
from the C6 atom whereas the N atom of the amine leaving group is at $1.60$ {\AA}.

@@

The formation of the Meissenheimer complex like structure with 
d$_{\text{NC}}=1.60$ {\AA} and d$_{\text{OC}}=1.48$ {\AA} is observed.}

@@

\caption{The free energy in $\text{Kcal}\;\text{mol}^{-1}$ as a function of the reaction 
parameter for the ADA catalyzed reaction. The Meissenheimer complex where the OH$^{-}$ anion
is bonded to the C atom and the leaving NH$_2$ moeity is also still bonded.}

@@

The free energy barrier is calculated to be $23\;\text{Kcal}\;\text{mol}^{-1}$

@@

free energy profile is to be obtained is partitioned into windows similar to 
umbrella sampling. But instead of using bias potentials to sample the configurations

@@

Using this ensemble of trajectories unbiased equilibrium probability distribution 

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Meta-data block
%% ---------------
%% Each author should be given as a separate \author command.
%%
%% Corresponding authors should have an e-mail given after the author
%% name as an \email command. Phone and fax numbers can be given
%% using \phone and \fax, respectively; this information is optional.
%%
%% The affiliation of authors is given after the authors; each
%% \affiliation command applies to all preceding authors not already
%% assigned an affiliation.
%%
%% The affiliation takes an option argument for the short name.  This
%% will typically be something like "University of Somewhere".
%%
%% The \altaffiliation macro should be used for new address, etc.
%% On the other hand, \alsoaffiliation is used on a per author basis
%% when authors are associated with multiple institutions.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\author{Sree Ganesh Balasubramani}
\affiliation{Department of Chemistry and Biochemistry, University of Arizona, Tucson, Arizona 85721, United States}
\author{Steven D. Schwartz}
\affiliation{Department of Chemistry and Biochemistry, University of Arizona, Tucson, Arizona 85721, United States}
%\author{I. Ken Groupleader}
%\altaffiliation{A shared footnote}
\email{sschwartz@email.arizona.edu}
%\phone{+123 (0)123 4445556}
%\fax{+123 (0)123 4445557}
%\affiliation[Unknown University]
%{Department of Chemistry, Unknown University, Unknown Town}
%\alsoaffiliation[Second University]
%{Department of Chemistry, Second University, Nearby Town}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% The document title should be given as usual. Some journals require
%% a running title from the author: this should be supplied as an
%% optional argument to \title.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\title[]
  {Transition path sampling based calculations of free energies for enzymatic
  reactions: the case of human methionine adenosyl transferase and plasmodium 
  vivax adenosine deaminase}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Some journals require a list of abbreviations or keywords to be
%% supplied. These should be set up here, and will be printed after
%% the title and author information, if needed.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\abbreviations{TPS,WHAM}
\keywords{American Chemical Society, \LaTeX}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% The abstract environment will automatically gobble the contents
%% if an abstract is not used by the target journal.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{abstract}
  Transition path sampling (TPS) is widely used for the
  calculations of reaction rates, transition state structures and reaction 
  coordinates of condensed phase systems. Here we discuss a scheme 
  for the calculation of free energies using the ensemble of TPS reactive 
  trajectories in combination with a window based sampling technique 
  for enzyme catalyzed reactions. We calculate the free energy profiles of 
  the reactions catalyzed by the human methionine S-adenosyltransferase (MAT2A) 
  enzyme and the plasmodium vivax adenosine deaminase ($pv$ADA) enzyme to assess 
  the accuracy of this method. MAT2A catalyzes the formation of S-adenosine-L-methionine
  following a S$_{\text{N}}2$ mechanism and using our method we estimate
  the free energy barrier for this reaction to 
  be $16\;\text{Kcal}\;\text{mol}^{-1}$ which
  is in excellent agreement with the experimentally measured activation 
  energy of $17.27\;\text{Kcal}\;\text{mol}^{-1}$. Furthermore, for the 
  $pv$ADA enzyme catalyzed reaction we estimate a free energy barrier of 
  $23\;\text{Kcal}\;\text{mol}^{-1}$ and the calculated free energy profile is
  similar to that predicted from experimental observations.
  Calculating free energies employing our simple method within TPS 
  provides significant advantages over such methods as umbrella sampling 
  since it is free from any applied external bias, accurate compared to 
  experimental measurements and has a reasonable computational cost.    
\end{abstract}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Start the main part of the manuscript here.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Introduction}
Molecular dynamics (MD) simulations are increasingly being
used to calculate quantitative estimates of experimentally measurable 
kinetic and thermodynamic parameters of enzyme catalyzed reactions. 
\cite{Karplus02NatStructMolBiol9p646,
Schramm98AnnuRevBiochem67p693,Zhang05AccChemRes38p379,
Schramm11AnnuRevBiochem80p703,Schramm18ChemRev118p11194} 
This has helped in proposing mechanisms and explanations for 
enzyme activity, fast high throughput screening of thousands of inhibitor 
molecules in drug lead discovery, \cite{Jorgensen09AccChemRes42p724,Sliwoski14PharmacolRev66p334} 
directed evolution for designing enzymes with enhanced catalytic 
activities, \cite{Thyme09Nature461p1300,Bloom09PNAS106p9995,Schafer19JAmChemSoc141p10431} etc.    
Straightforward MD simulations adequately sample 
the long-lived stable reactant and product states 
but the high energy barrier-crossing events that occur rarely and at long 
time scales are not easily accessible using this method alone.
To access such states at a higher frequency using MD simulations 
several enhanced sampling techniques 
have been developed in the past few decades. These include 
umbrella sampling, \cite{Kastner11WileyInterdiscipRevComputMolSci1p932} 
metadynamics, \cite{Barducci11WileyInterdiscipRevComputMolSci1p826}
steered MD, \cite{Park04JChemPhys120p5946}
milestoning, \cite{Faradjian04JChemPhys120p10880} and transition path sampling 
(TPS). \cite{Pratt86JChemPhys85p5045,Bolhuis02AnnRevPhysChem53p291,
Dellago98JChemPhys108p1964,Bolhuis21AdvTheory4p2000237}

TPS is an attractive method since it samples reactive trajectories
in which the rare barrier-crossing events are guaranteed to occur and
it does not require a priori knowledge of the reaction coordinate which can be 
complex for large biological systems.
TPS is based on Monte Carlo importance sampling in the 
space of trajectories. It uses the shooting and shifting algorithm 
to collect an ensemble of reactive pathways that are real dynamical 
trajectories generated without the application of any bias forces. \cite{dellago02AdvChemPhys123} 
Therefore the TPS ensemble can be used to obtain 
kinetic information such as the rate of the reaction as well as to elucidate 
the transition state, reaction coordinate and the reaction mechanism. 
TPS has been successfully applied to study problems such as enzyme catalysis, 
\cite{Antoniou01JPhysChemB105p5553,Hay12NatChem4p161,Schwartz09NatChemBiol5p551} 
and protein folding. \cite{Bolhuis03ProcNatlAcadSci100p12129,Juraszek12ChemPhys396p30}

Calculating equilibrium properties such as the free energy of an enzymatic 
catalysis reaction within TPS is not straightforward. 
The Monte Carlo algorithm used in TPS has an acceptance criterion that 
is only satisfied if the trajectory begins in the reactant basin and ends in 
the product basin or vice versa, hence trajectories that are localized in the 
reactant or product basins, without visiting both are rejected. 
If one defines an order parameter $\zeta$, which separates reactant from 
product basins, then all trajectories, reactive and non-reactive, 
contribute to the probability density distribution function 
($P(\zeta)$) which is the probability of a system at equilibrium to have 
a particular value of the order parameter. \cite{Dellago09AdvCompSimAppp167}

Frenkel has shown that inclusion of rejected states in a Markov state Monte 
Carlo algorithm can help speed up the convergence of the calculations of 
the order parameter distribution of a two-dimensional
Ising model. \cite{Frenkel04ProcNatAcadSci101p17571}
Inclusion of rejected trajectories within TPS was first discussed by Radhakrishnan 
et al. \cite{Radhakrishnan04JChemPhys121p2436} who developed `BOLAS' which combines 
the shooting and shifting algorithms of TPS with a window based sampling technique 
in the spirit of umbrella sampling to calculate equilibrium free energy 
differences. Peters et al. used a variant of the BOLAS scheme called the 
equilibrium path sampling (EPS) where they use aimless shooting 
combined with window based sampling for free energy calculations. 
\cite{Peters08JAmChemSoc130p17342,Beckham10epsbook} 
More recently Brotzakis et al., \cite{Brotzakis19JChemPhys151p174111}
have developed a re-weighted path ensemble scheme within the standard TPS method 
called virtual interface exchange TPS (VIE-TPS) to calculate the free energy 
landscape. VIE-TPS is shown to produce reasonably accurate free energy profiles for 
chemical reactions that are diffusive in nature whereas it is less accurate otherwise.
Transition interface sampling (TIS) has been used to obtain 
a re-weighted path ensemble which can give access to free energies 
and committor surfaces. \cite{Rogal10JChemPhys17p174109}

@@

distributed to almost all body tissues and fluids, is a universal 

@@

Plasmodium vivax is a parasite that is responsible for the largest number 

@@

is a possible anti-malarial drug development strategy. \cite{Madrid08JBiolChem283p35899} 

@@

to calculate the free energies, then we show calculations which apply this 
method on the two enzymatic reactions. Finally we compare our results to experiments

@@

introducing notations as well as for the sake of completeness. 

@@

on $z_{i\Delta t}$ and not on conditions before the time $i\Delta t$ the total path probability can 

@@

Within the transition path sampling, the trajectories of interest start in the reactant region of the 
phase space ($\mathcal{A}$) and end in the product region ($\mathcal{B}$). Such reactive trajectories have 
a restricted probability distribution function given by
\begin{equation}
\mathcal{P}_{\mathcal{AB}}[z_{\mathcal{T}}] = h_{\mathcal{A}}(z_0)\mathcal{P}[z(\mathcal{T})]
h_{\mathcal{B}}(z_{\mathcal{T}})/Z_{\mathcal{AB}}(\mathcal{T})\label{eqn:tpsensem}
\end{equation}
where 
\[
    h_{\mathcal{A}/\mathcal{B}}(z)= 
\begin{cases}
    1, & \text{if } z\in \mathcal{A}/\mathcal{B}\\
    0,              & \text{otherwise}
\end{cases}
\]
and $Z_{\mathcal{AB}}(\mathcal{T})$ is the normalization factor for this 
probability distribution function.
The set of all reactive trajectories characterized by the probability 
distribution function given by Eq. \ref{eqn:tpsensem} is the 
transition path ensemble.

@@

The free energy of a molecular system as a function of the order parameter $\zeta$ 
is defined as 

@@

where $P(\zeta)$ is the probability distribution function of the reaction coordinate
$\zeta$, $C$ is an arbitrary constant, $k_{\text{B}}$ is the Boltzmann constant
and $T$ is the temperature. At equilibrium the probability distribution of the 
reaction coordinate can be obtained from the distribution function for the phase 
space $\rho(\textbf{q})$ as

@@

The integration is over the entire phase space. In practice, the calculation of this distribution function

@@

divided into bins and the frequency of the occurrence of the reaction coordinate 
within a particular window during the course of MD simulation is used to calculate 
the probability distribution function. Since some values of the reaction coordinates 
particularly close to the transition state are 

@@

without the application of any external bias, but the trajectories 
sampled with TPS are only reactive and hence

@@

The restriction that the pathways start from the reactant state and end in the 
product state can be relaxed to obtain an ensemble of trajectories that will 
be distributed according to the equilibrium distribution. 

@@

\item The order parameter is divided into windows $\{\zeta_i\}$ and within each window,
$\zeta_{i}^{min} < \zeta < \zeta_{i}^{max}$ is to be sampled.

@@

$\tilde{z}(\mathcal{T})$ and accept the new trajectory if for any time slice the calculated 

@@

accepted trajectories for the window referenced by the index $i$. 
\item Calculate the free energies for each of the windows according to Eq. \ref{eqn:fenergy}
and combine them such that the function $\text{A}(\zeta)$ is continuous 
by adjusting the constants $C$. 

@@

The energies of both the systems were minimized using 50 steps of the 
steepest descent method, followed by 2000 steps of the
adopted basis Newton-Raphson method where only classical molecular mechanics 
was used for the dynamics. 
The minimized systems were
heated slowly to 300 K for 35 ps beginning with harmonic
constraints on all atoms except on the H atoms and the TIP3P 
water molecules with gradual reduction of the restraint forces. 
15 ps of equilibration was carried out starting with harmonic 
restraint forces followed by 20 ps of constraint free 
equilibration to prepare the systems for TPS simulations. 
During the heating and equilibration steps, bonds that contain 
hydrogen atoms in the MM region were restrained to their equilibrium values 
using the SHAKE procedure. \cite{Ryckaert77JComputPhys23p327} 
%------------------------------------------------------------------------------ 
\subsection{Transition path sampling}
%------------------------------------------------------------------------------
TPS simulations require the definition of unique reactant and product states
to collect reactive trajectories based on the Monte Carlo algorithm. Here we 
discuss the details of the reactions involving the MAT2A and the $pv$ADA enzymes 
to identify and define the order parameters characterizing the reactant and product basins.

The reaction catalyzed by the human MAT2A enzyme is
between adenosyl $5'$-triphosphate (ATP) and L-methionine (MET) molecules 
resulting in the formation of S-adenosyl-L-methionine (SAM) which is 
depicted in Fig. \ref{fig:mat2a-reaction}. This reaction follows a S$_{\text{N}}2$ 
mechanism and was experimentally characterized by 
Firestone et al. \cite{Firestone17JAmChemSoc139p13754}. The key bond 
parameters for this reaction consists of the distance between the
nucleophilic sulfur atom of MET and the $5'-C$ atom of the ATP (denoted as d$_{\mathrm{SC}}$), 
and the distance between oxygen atom of the phosphate group and the $5'-C$ 
atom of ATP (d$_{\mathrm{OC}}$). For the TPS calculations we define the 
reactant state to have d$_{\mathrm{OC}}$ to be less than 1.7 {\AA} and the product 
state to have the d$_{\mathrm{SC}}$ less than 2.0 {\AA}.

\begin{figure}
\includegraphics[width=\columnwidth]{figures/mat2a-reaction.png}
\caption{The reaction catalyzed by the human MAT2A enzyme between ATP and L-methionine resulting in the 
formation of S-adenosyl L-methionine and triphosphate leaving group.}
\label{fig:mat2a-reaction}
\end{figure}

$pv$ADA catalyzes the reaction between adenosine and OH$^{-}$ ion 
resulting in the formation of inosine and ammonia as shown in
Fig. \ref{fig:ada-reaction}.
The nucleophilic O atom of the OH$^{-}$ anion
attacks the electrophilic C6 atom of the adenosyl ring leading to the 
formation of a Meisenheimer complex like structure while the N1 atom 
on the adenosyl moiety gets protonated by the Glu229 residue of $pv$ADA. 
The key bond parameters are the distance between the approaching hydroxyl 
oxygen and the C6 atom (denoted as d$_{\text{OC}}$) and the 
dissociating C6$-$N6 (d$_{\text{NC}}$) bond distance. For the TPS simulations 
we define the reactant state to have d$_{\text{NC}}$ being less than $1.8$ {\AA} 
and the product state to have the d$_{\text{OC}}$ less than $1.8$ {\AA}.

\begin{figure}
\includegraphics[width=\columnwidth]{figures/ada-new-reaction.png}
\caption{The reaction catalyzed by the $pv$ADA enzyme between adenosine and hydroxide ion
resulting in the formation of inosine and ammonia.}
\label{fig:ada-reaction}
\end{figure}

@@

250 steps forwards and backwards using $1\;fs$ time steps

@@

The order parameter for the MAT2A system 

@@

is used to guide the window based umbrella sampling simulations (though in this case,
of course, this is an unbiased dynamically exact reactive trajectory). 
To sample in a particular window $\zeta_i$, a frame from this reactive 

@@

order parameter ($\zeta$) falls within 
$\zeta_{i}^{min} < \zeta < \zeta_{i}^{max}$. Using the shooting algorithm

@@

\caption{Equilibrated structure of the MAT2A enzyme in complex with ATP and 

@@

repeating this procedure many times and finding the fraction of trajectories that go 

@@

promote the reaction. \cite{Schramm18Biochem57p3299} Taking this into account the 
committor distribution analysis
begins by constraining the residues and molecules that are hypothesized to be part of the 

@@

\text{Kcal mol}^{-1}$.

@@

\caption{The free energy as a function of the order 
parameter for the reaction catalyzed
by the human MAT2A enzyme. The activation barrier is 
calculated to be $16\;\text{Kcal mol}^{-1}$. The solid colored disks represent the free energies 
calculated within the individual windows which is adjusted to make the free energy as a function 
of the order parameter a continuous function.}

@@

\caption{Equilibrated structure of $pv$ADA enzyme complexed with adenosine, 
OH$^{-}$ anion and Zn$^{2+}$.}

@@

The approaching O atom of the OH$^{-}$ anion is at a distance of $1.49$ {\AA}
from the C6 atom whereas the N atom of the amine leaving group is at $1.62$ {\AA}
from the C6 atom.

@@

The formation of the Meisenheimer complex like structure with 
d$_{\text{NC}}=1.62$ {\AA} and d$_{\text{OC}}=1.49$ {\AA} is observed.}

@@

\caption{The free energy as a function of the reaction 
parameter for the ADA catalyzed reaction. The activation barrier is calculated as 
$21\;\text{Kcal}\;\text{mol}^{-1}$. The solid colored disks represent the free energies 
calculated within the individual windows which is adjusted to make the free energy as a function 
of the order parameter a continuous function.}

@@

The free energy barrier is calculated to be $21\;\text{Kcal}\;\text{mol}^{-1}$

@@

free energy profile is to be obtained is partitioned into windows as is done in 
umbrella sampling, but instead of using bias potentials to sample the configurations

@@

Using this ensemble of trajectories the unbiased equilibrium probability distribution 
